---
title: "Lecture 4: Doubly Robust Estimation"
author: "Se Yoon Lee"
date: "2026-06-27"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: readable
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r picture-helpers, include=FALSE}
start_picture <- function() {
  par(mar = c(0, 0, 0, 0), xaxs = "i", yaxs = "i")
  plot.new()
  plot.window(xlim = c(0, 1), ylim = c(0, 1))
}

draw_text <- function(x, y, label, cex = 1, font = 1, col = "#222222") {
  lines <- strsplit(label, "\n", fixed = TRUE)[[1]]
  line_height <- 0.055 * cex
  offsets <- seq((length(lines) - 1) / 2, -(length(lines) - 1) / 2) *
    line_height
  text(rep(x, length(lines)), y + offsets, labels = lines,
       cex = cex, font = font, col = col)
}

draw_box <- function(x, y, w, h, label, fill = "#FFFFFF",
                     border = "#333333", lwd = 2, lty = 1,
                     cex = 1, font = 1) {
  rect(x - w / 2, y - h / 2, x + w / 2, y + h / 2,
       col = fill, border = border, lwd = lwd, lty = lty)
  draw_text(x, y, label, cex = cex, font = font)
}

draw_circle <- function(x, y, r, label, fill = "#FFFFFF",
                        border = "#333333", lwd = 2, cex = 1,
                        font = 1) {
  symbols(x, y, circles = r, inches = FALSE, add = TRUE,
          bg = fill, fg = border, lwd = lwd)
  draw_text(x, y, label, cex = cex, font = font)
}

draw_arrow <- function(x0, y0, x1, y1, col = "#333333", lwd = 2) {
  arrows(x0, y0, x1, y1, length = 0.08, lwd = lwd, col = col)
}
```

## Today's Question

Lecture 2 used **inverse probability weighting**:

> Model the treatment: who is likely to be a smoker, based on baseline covariates?

Lecture 3 used **outcome regression**:

> Model the outcome: who is likely to have CVD, based on smoking and baseline covariates?

Lecture 4 combines both ideas:

> Can we use both the treatment model and the outcome model together?

This is called **doubly robust estimation**.

Our treatment, outcome, and covariates are the same as before:

| Symbol | Meaning in this lecture |
|:--|:--|
| `T` | Smoking status: 1 = smoker, 0 = non-smoker |
| `Y` | CVD status: 1 = had CVD, 0 = did not have CVD |
| `X` | Age, gender, race, education, income-to-poverty ratio, and BMI |

As before, this teaching analysis does **not use survey weights**.

## The Big Picture

Doubly robust estimation uses two helpers.

```{r two-models-picture, echo=FALSE, fig.width=8.8, fig.height=4.8}
start_picture()

draw_box(0.11, 0.68, 0.18, 0.18, "Baseline\ncovariates\nX",
         fill = "#F1CE63", cex = 0.85, font = 2)
draw_box(0.34, 0.80, 0.22, 0.15, "Treatment model\nP(T = 1 | X)",
         fill = "#A0CBE8", cex = 0.78, font = 2)
draw_box(0.34, 0.48, 0.22, 0.15, "Outcome model\nE(Y | T, X)",
         fill = "#FCE4C6", cex = 0.78, font = 2)
draw_box(0.61, 0.64, 0.20, 0.20, "Doubly robust\nestimate",
         fill = "#B7D7A8", cex = 0.9, font = 2)
draw_box(0.86, 0.64, 0.14, 0.18, "Adjusted\nCVD risk\ndifference",
         fill = "#E8E8E8", cex = 0.72, font = 2)

draw_arrow(0.20, 0.72, 0.23, 0.79)
draw_arrow(0.20, 0.64, 0.23, 0.50)
draw_arrow(0.45, 0.78, 0.51, 0.69, col = "#4E79A7", lwd = 3)
draw_arrow(0.45, 0.50, 0.51, 0.59, col = "#F28E2B", lwd = 3)
draw_arrow(0.71, 0.64, 0.79, 0.64)

draw_text(0.50, 0.18, "Treatment model + outcome model + correction from observed data",
          cex = 0.95, font = 2)
```

The name "doubly robust" comes from a powerful idea:

> If either the treatment model is good enough or the outcome model is good enough, the estimator can still work well.

But it is not magic. If both models are badly wrong, the estimate can still be wrong.

## Why Add a Correction?

Outcome regression predicts risk under `T = 0` and under `T = 1`. Doubly robust estimation starts there, then adds a correction using the observed outcomes.

```{r correction-picture, echo=FALSE, fig.width=8.8, fig.height=4.5}
start_picture()

draw_box(0.18, 0.62, 0.24, 0.20, "Model prediction\nunder T = 1\nor T = 0",
         fill = "#FCE4C6", cex = 0.82, font = 2)
draw_box(0.48, 0.62, 0.24, 0.20, "Correction from\nobserved data",
         fill = "#A0CBE8", cex = 0.82, font = 2)
draw_box(0.78, 0.62, 0.24, 0.20, "Doubly robust\nrisk estimate",
         fill = "#B7D7A8", cex = 0.82, font = 2)

draw_text(0.33, 0.62, "+", cex = 2.0, font = 2)
draw_text(0.63, 0.62, "=", cex = 2.0, font = 2)

draw_text(
  0.50,
  0.24,
  "The correction checks whether the model prediction\nmatches what we actually observed.",
  cex = 0.92,
  font = 2
)
```

If the outcome model predicts very well, the correction is small. If the outcome model misses something but the treatment model is good, the correction can help repair the estimate.

## Load and Prepare the Data

```{r load-data}
data_path <- file.path("..", "Data", "NHANES_data.csv")

if (!file.exists(data_path)) {
  data_path <- file.path("Data", "NHANES_data.csv")
}

nhanes <- read.csv(data_path, stringsAsFactors = FALSE)

baseline_vars <- c(
  "age_yr",
  "gender",
  "race",
  "educ_lvl",
  "inc_to_pov_ratio",
  "bmi"
)

analysis_vars <- c("smoker_indicator", "cvd_indicator", baseline_vars)

dr_data <- nhanes[complete.cases(nhanes[, analysis_vars]), analysis_vars]

dr_data$smoker_indicator <- as.integer(dr_data$smoker_indicator)
dr_data$cvd_indicator <- as.integer(dr_data$cvd_indicator)

dr_data$smoking_group <- ifelse(
  dr_data$smoker_indicator == 1,
  "Smoker",
  "Non-smoker"
)

dr_data$smoking_group <- factor(
  dr_data$smoking_group,
  levels = c("Non-smoker", "Smoker")
)

dr_data$gender_label <- factor(
  dr_data$gender,
  levels = c("F", "M"),
  labels = c("Female", "Male")
)

dr_data$race_label <- factor(
  dr_data$race,
  levels = c(
    "mex_american",
    "other_hispanic",
    "nh_white",
    "nh_black",
    "nh_asian"
  ),
  labels = c(
    "Mexican American",
    "Other Hispanic",
    "Non-Hispanic White",
    "Non-Hispanic Black",
    "Non-Hispanic Asian"
  )
)

dr_data$educ_label <- factor(
  dr_data$educ_lvl,
  levels = c(
    "lt_9th_grade",
    "9_to_12th_grade_no_diploma",
    "hs_grad_or_ged",
    "start_college_to_aa",
    "college_grad_or_above"
  ),
  labels = c(
    "Less than 9th grade",
    "9th-12th, no diploma",
    "High school or GED",
    "Some college or AA",
    "College graduate or above"
  )
)

nrow(dr_data)
```

We use the same complete-case sample as Lectures 2 and 3.

```{r sample-summary}
sample_summary <- data.frame(
  Quantity = c(
    "Number of people",
    "Number of smokers",
    "Number of non-smokers",
    "Number with CVD"
  ),
  Value = c(
    nrow(dr_data),
    sum(dr_data$smoker_indicator == 1),
    sum(dr_data$smoker_indicator == 0),
    sum(dr_data$cvd_indicator == 1)
  )
)

knitr::kable(sample_summary)
```

## Step 1: Fit the Treatment Model

The treatment model estimates the propensity score:

```{r propensity-formula, echo=FALSE, results='asis'}
cat("Propensity score = P(smoker | age, gender, race, education, income, BMI)")
```

```{r treatment-model}
treatment_model <- glm(
  smoker_indicator ~ age_yr + gender_label + race_label +
    educ_label + inc_to_pov_ratio + bmi,
  data = dr_data,
  family = binomial()
)

dr_data$propensity_score <- predict(treatment_model, type = "response")

summary(dr_data$propensity_score)
```

```{r propensity-picture, echo=FALSE, fig.width=7.5, fig.height=4.3}
old_par <- par(mar = c(5, 4.5, 4.5, 1))

boxplot(
  propensity_score ~ smoking_group,
  data = dr_data,
  col = c("#A0CBE8", "#F28E2B"),
  ylab = "Estimated probability of being a smoker",
  main = ""
)

mtext(
  "Treatment Model: Propensity Scores",
  side = 3,
  line = 1.3,
  font = 2,
  cex = 1.1
)

par(old_par)
```

## Step 2: Fit the Outcome Model

The outcome model estimates CVD risk:

```{r outcome-formula, echo=FALSE, results='asis'}
cat("Outcome model = E(CVD | smoking, age, gender, race, education, income, BMI)")
```

```{r outcome-model}
outcome_model <- glm(
  cvd_indicator ~ smoker_indicator + age_yr + gender_label +
    race_label + educ_label + inc_to_pov_ratio + bmi,
  data = dr_data,
  family = binomial()
)

dr_data$predicted_risk_observed <- predict(outcome_model, type = "response")

data_if_non_smoker <- dr_data
data_if_smoker <- dr_data

data_if_non_smoker$smoker_indicator <- 0
data_if_smoker$smoker_indicator <- 1

dr_data$predicted_risk_if_non_smoker <- predict(
  outcome_model,
  newdata = data_if_non_smoker,
  type = "response"
)

dr_data$predicted_risk_if_smoker <- predict(
  outcome_model,
  newdata = data_if_smoker,
  type = "response"
)

outcome_model_summary <- data.frame(
  Quantity = c(
    "Average predicted risk if everyone non-smoker",
    "Average predicted risk if everyone smoker"
  ),
  Percent = round(
    100 * c(
      mean(dr_data$predicted_risk_if_non_smoker),
      mean(dr_data$predicted_risk_if_smoker)
    ),
    1
  )
)

knitr::kable(outcome_model_summary)
```

```{r outcome-model-picture, echo=FALSE, fig.width=7.5, fig.height=4.3}
old_par <- par(mar = c(5, 4.5, 4.5, 1))

boxplot(
  predicted_risk_observed ~ smoking_group,
  data = dr_data,
  col = c("#A0CBE8", "#F28E2B"),
  ylab = "Predicted CVD risk",
  main = ""
)

mtext(
  "Outcome Model: Predicted CVD Risk",
  side = 3,
  line = 1.3,
  font = 2,
  cex = 1.1
)

par(old_par)
```

## Step 3: Build the Doubly Robust Estimate

The doubly robust estimator starts with the outcome-regression prediction and adds a correction.

For the smoker world, `T = 1`:

```{r dr-formula-smoker, echo=FALSE, results='asis'}
cat("DR risk under T = 1 = average(predicted risk if T = 1 + smoker correction)")
```

For the non-smoker world, `T = 0`:

```{r dr-formula-nonsmoker, echo=FALSE, results='asis'}
cat("DR risk under T = 0 = average(predicted risk if T = 0 + non-smoker correction)")
```

The correction uses the observed outcome minus the model prediction. That difference is called a **residual**.

```{r doubly-robust-estimator}
T_obs <- dr_data$smoker_indicator
Y_obs <- dr_data$cvd_indicator
e_hat <- dr_data$propensity_score
m1_hat <- dr_data$predicted_risk_if_smoker
m0_hat <- dr_data$predicted_risk_if_non_smoker

dr_data$correction_if_smoker <- T_obs / e_hat * (Y_obs - m1_hat)
dr_data$correction_if_non_smoker <- (1 - T_obs) / (1 - e_hat) *
  (Y_obs - m0_hat)

dr_risk_smoker <- mean(m1_hat + dr_data$correction_if_smoker)
dr_risk_non_smoker <- mean(m0_hat + dr_data$correction_if_non_smoker)

dr_difference <- dr_risk_smoker - dr_risk_non_smoker
dr_ratio <- dr_risk_smoker / dr_risk_non_smoker

dr_components <- data.frame(
  Quantity = c(
    "Outcome-model risk if everyone non-smoker",
    "Average correction for non-smoker world",
    "Doubly robust risk if everyone non-smoker",
    "Outcome-model risk if everyone smoker",
    "Average correction for smoker world",
    "Doubly robust risk if everyone smoker"
  ),
  Percent = round(
    100 * c(
      mean(m0_hat),
      mean(dr_data$correction_if_non_smoker),
      dr_risk_non_smoker,
      mean(m1_hat),
      mean(dr_data$correction_if_smoker),
      dr_risk_smoker
    ),
    2
  )
)

knitr::kable(dr_components)
```

In this dataset, the correction is small. That means the IPW and outcome-regression pieces are giving fairly similar answers.

## Picture of the Doubly Robust Calculation

```{r dr-components-picture, echo=FALSE, fig.width=8.5, fig.height=4.8}
component_matrix <- rbind(
  c(100 * mean(m0_hat), 100 * mean(m1_hat)),
  c(100 * mean(dr_data$correction_if_non_smoker),
    100 * mean(dr_data$correction_if_smoker)),
  c(100 * dr_risk_non_smoker, 100 * dr_risk_smoker)
)

colnames(component_matrix) <- c("T = 0", "T = 1")
rownames(component_matrix) <- c("Outcome model", "Correction", "DR estimate")

old_par <- par(mar = c(5, 4.5, 4.8, 1))

bar_locations <- barplot(
  component_matrix,
  beside = TRUE,
  col = c("#FCE4C6", "#A0CBE8", "#B7D7A8"),
  ylim = c(
    min(0, min(component_matrix) * 1.4),
    max(component_matrix) * 1.25
  ),
  ylab = "Percent with CVD",
  main = "",
  cex.names = 0.95
)

mtext(
  "Doubly Robust Risk = Prediction + Correction",
  side = 3,
  line = 1.4,
  font = 2,
  cex = 1.05
)

text(
  x = bar_locations,
  y = component_matrix + ifelse(component_matrix >= 0, 0.45, -0.45),
  labels = paste0(round(component_matrix, 2), "%"),
  cex = 0.75
)

legend(
  "topright",
  legend = rownames(component_matrix),
  fill = c("#FCE4C6", "#A0CBE8", "#B7D7A8"),
  bty = "n"
)

par(old_par)
```

## Compare All Four Methods

Now we compare:

1. Simple observed comparison
2. IPW
3. Outcome regression
4. Doubly robust estimation

```{r compare-methods}
smoker_rows <- dr_data$smoking_group == "Smoker"
non_smoker_rows <- dr_data$smoking_group == "Non-smoker"

observed_risk_smoker <- mean(dr_data$cvd_indicator[smoker_rows])
observed_risk_non_smoker <- mean(dr_data$cvd_indicator[non_smoker_rows])
observed_difference <- observed_risk_smoker - observed_risk_non_smoker
observed_ratio <- observed_risk_smoker / observed_risk_non_smoker

ipw_weight <- ifelse(
  dr_data$smoker_indicator == 1,
  1 / dr_data$propensity_score,
  1 / (1 - dr_data$propensity_score)
)

weighted_risk <- function(outcome, weights) {
  sum(weights * outcome) / sum(weights)
}

ipw_risk_smoker <- weighted_risk(
  dr_data$cvd_indicator[smoker_rows],
  ipw_weight[smoker_rows]
)

ipw_risk_non_smoker <- weighted_risk(
  dr_data$cvd_indicator[non_smoker_rows],
  ipw_weight[non_smoker_rows]
)

ipw_difference <- ipw_risk_smoker - ipw_risk_non_smoker
ipw_ratio <- ipw_risk_smoker / ipw_risk_non_smoker

or_risk_smoker <- mean(dr_data$predicted_risk_if_smoker)
or_risk_non_smoker <- mean(dr_data$predicted_risk_if_non_smoker)
or_difference <- or_risk_smoker - or_risk_non_smoker
or_ratio <- or_risk_smoker / or_risk_non_smoker

comparison_table <- data.frame(
  Method = c(
    "Observed comparison",
    "IPW",
    "Outcome regression",
    "Doubly robust"
  ),
  Non_smoker_risk_percent = round(
    100 * c(
      observed_risk_non_smoker,
      ipw_risk_non_smoker,
      or_risk_non_smoker,
      dr_risk_non_smoker
    ),
    1
  ),
  Smoker_risk_percent = round(
    100 * c(
      observed_risk_smoker,
      ipw_risk_smoker,
      or_risk_smoker,
      dr_risk_smoker
    ),
    1
  ),
  Difference_percentage_points = round(
    100 * c(
      observed_difference,
      ipw_difference,
      or_difference,
      dr_difference
    ),
    1
  ),
  Risk_ratio = round(c(observed_ratio, ipw_ratio, or_ratio, dr_ratio), 2)
)

knitr::kable(
  comparison_table,
  col.names = c(
    "Method",
    "Non-smoker risk (%)",
    "Smoker risk (%)",
    "Difference (percentage points)",
    "Risk ratio"
  )
)
```

```{r compare-methods-picture, echo=FALSE, fig.width=8.5, fig.height=4.8}
method_differences <- comparison_table$Difference_percentage_points
names(method_differences) <- c("Observed", "IPW", "Outcome\nregression", "Doubly\nrobust")

old_par <- par(mar = c(6, 7, 5, 1))

bar_locations <- barplot(
  method_differences,
  col = c("#A0CBE8", "#F28E2B", "#59A14F", "#B7D7A8"),
  ylim = c(0, max(method_differences) * 1.35),
  ylab = "Difference in CVD risk\npercentage points",
  main = "",
  cex.names = 0.9
)

mtext(
  "Estimated Smoker Minus Non-Smoker Risk Difference",
  side = 3,
  line = 1.6,
  font = 2,
  cex = 1.05
)

text(
  x = bar_locations,
  y = method_differences + 0.55,
  labels = paste0(method_differences, " pp"),
  cex = 0.9
)

par(old_par)
```

The doubly robust estimate is close to the IPW and outcome-regression estimates in this example. That is reassuring because two different adjustment strategies tell a similar story.

## Interpretation

The simple observed comparison estimates a CVD risk difference of `r round(100 * observed_difference, 1)` percentage points.

After adjustment:

| Method | Adjusted risk difference |
|:--|:--|
| IPW | `r round(100 * ipw_difference, 1)` percentage points |
| Outcome regression | `r round(100 * or_difference, 1)` percentage points |
| Doubly robust | `r round(100 * dr_difference, 1)` percentage points |

The doubly robust estimate says:

> In this complete-case classroom sample, the adjusted predicted CVD risk is `r round(100 * dr_risk_non_smoker, 1)`% if everyone were set to non-smoker and `r round(100 * dr_risk_smoker, 1)`% if everyone were set to smoker.

The adjusted difference is `r round(100 * dr_difference, 1)` percentage points.

This is smaller than the raw observed difference. That suggests that some of the raw smoker versus non-smoker gap was related to baseline covariate differences between the groups.

## What Can Go Wrong?

Doubly robust does not mean "always correct."

```{r double-robustness-picture, echo=FALSE, fig.width=8.8, fig.height=5.2}
start_picture()

draw_box(0.18, 0.76, 0.26, 0.14, "Treatment model\ngood enough",
         fill = "#A0CBE8", cex = 0.85, font = 2)
draw_box(0.50, 0.76, 0.26, 0.14, "Outcome model\nmay be wrong",
         fill = "#F5F5F5", cex = 0.85, font = 2)
draw_box(0.82, 0.76, 0.22, 0.14, "DR estimate\ncan work well",
         fill = "#B7D7A8", cex = 0.82, font = 2)

draw_arrow(0.31, 0.76, 0.37, 0.76)
draw_arrow(0.63, 0.76, 0.71, 0.76)

draw_box(0.18, 0.50, 0.26, 0.14, "Treatment model\nmay be wrong",
         fill = "#F5F5F5", cex = 0.85, font = 2)
draw_box(0.50, 0.50, 0.26, 0.14, "Outcome model\ngood enough",
         fill = "#FCE4C6", cex = 0.85, font = 2)
draw_box(0.82, 0.50, 0.22, 0.14, "DR estimate\ncan work well",
         fill = "#B7D7A8", cex = 0.82, font = 2)

draw_arrow(0.31, 0.50, 0.37, 0.50)
draw_arrow(0.63, 0.50, 0.71, 0.50)

draw_box(0.18, 0.24, 0.26, 0.14, "Treatment model\nbad",
         fill = "#F5F5F5", cex = 0.85, font = 2)
draw_box(0.50, 0.24, 0.26, 0.14, "Outcome model\nbad",
         fill = "#F5F5F5", cex = 0.85, font = 2)
draw_box(0.82, 0.24, 0.22, 0.14, "Estimate may\nbe biased",
         fill = "#E15759", cex = 0.82, font = 2)

draw_arrow(0.31, 0.24, 0.37, 0.24)
draw_arrow(0.63, 0.24, 0.71, 0.24)

draw_text(0.50, 0.08, "At least one model must be reasonable. Both bad is trouble.",
          cex = 0.95, font = 2)
```

We still need important causal assumptions:

1. No important unmeasured confounders.
2. Good overlap: people with similar `X` values could plausibly be smokers or non-smokers.
3. At least one of the two models is reasonable.
4. The data variables are measured well enough for the teaching question.

## Key Takeaways

1. IPW uses a treatment model.
2. Outcome regression uses an outcome model.
3. Doubly robust estimation uses both models.
4. It starts with model-predicted risks and adds a correction from observed outcomes.
5. It can work well if at least one of the two models is reasonable, but it is not magic.
